---
title: "Flight Delays and Cancellations in the U.S."
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)

```

```{r, include=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(caret)
library(knitr)
library(ggplot2)
library(dplyr)
library(ggpattern)
library(tidyr)
library(readr)
library(viridis)     
library(patchwork)
library(pROC)
```

```{r}
flights <- read_csv(
  "/Users/chrysanthemum/Downloads/flight_data_2024_sample.csv",
  col_types = cols()
)
```

# Background and Objectives
Flight delays and cancellations remain a persistent challenge in the aviation industry, affecting millions of passengers annually and creating significant financial burdens for airlines. While these disruptions typically stem from routine factors—such as weather, security processing, carrier logistics, or scheduling issues—2025 has introduced unprecedented shocks that have exacerbated the situation.

Notably, the industry faced a record-breaking U.S. federal government shutdown. This crisis forced air traffic controllers and security staff to work without pay, resulting in widespread staffing shortages and operational paralysis at major airports. Furthermore, media coverage has intensified surrounding two catastrophic events: the June crash of an Air India flight from Ahmedabad to London, which sparked global debates on regulatory oversight, and the collision between an American Airlines jet and an army helicopter over the Potomac River. Severe disruptions like these have not only grounded flights but have shaken the public's trust in air travel safety. 

Inspired by these events, our project aims to examine how major disruptions can contribute to severe airline delays. For this project we analyze a comprehensive sample dataset of U.S. commercial flights in 2024, containing information about airlines, routes, timings, delays, cancellations and operational factors. The goal of our project is to analyze these delay and cancellation patters through exploratory data analysis in 3 different ways: an analysis of airline, airport and route delays & cancellations, a temporal analysis of delays & cancellations, and geographical patterns of these delays & cancellations. The goal is to identify significant patterns in delay behavior and visualize how delays and cancellations vary across airlines, airports, routes and cities. Following this, we use boosting and linear regression to build a model capable of predicting whether a flight is likely to be delayed. 

Following our analysis using supervised learning, we found that our predictive models achieved acceptable overall accuracy but exhibited poor specificity, largely because the models tended to classify nearly all flights as on-time. This behavior is driven by a pronounced class imbalance in the dataset, with substantially more on-time flights than delayed flights, making it difficult for the models to reliably identify the occurrence of delays. Accordingly, we model two related but distinct outcomes: the occurrence of flight delays and the length of departure delays. For delay occurrence, we initially apply a boosting classification model, but due to its limited discriminative performance, we transitioned to unsupervised learning methods to better diagnose the structural reasons underlying this poor prediction. In contrast, when modeling delay length, a linear regression model reveals that travel season and departure hour are strong and statistically significant predictors, indicating that while supervised models struggle to identify whether delays occur, they are more effective at explaining the severity of delays once they arise.

Due to time constraints and our project being a pilot-level study, we were unable to address variables that we thought would also affect delays and cancellations. To improve future research, we would recommend incorporating datasets and variables that relate to weather conditions, airport congestion, number of active flights, etc. This could further build on and improve the model that we have explored in this project, providing a more comprehensive analysis of the cause of flight delays in the United States.

# Methodology

**Data Used**

We used a sample dataset of all commercial flights at U.S. airports in 2024 because we wanted a complete, single-year snapshot of aviation activity. The dataset, which we found on Kaggle, contains 10,000 rows of detailed flight-level observations, including airline identifiers, airport codes, cities, and both arrival and departure delay times, indicators for whether the flight was cancelled or not, along with the categorized reasons for those delays.

To visualize these patterns geographically, we required the latitude and longitude of each arrival and departure airport. For this, we used a second Kaggle dataset that provides geographic coordinates and metadata for all commercial airports in the United States. We filtered this dataset to retain only the airports appearing in our 2024 flight sample, ensuring that each flight could be accurately plotted on a U.S. map. 

We also used a dataset containing the the year, month, and # of passengers travelling domestically and internationally. This dataset was used to track seasonal travelling patterns across years, only focusing on the domestic numbers (since we are only looking at flights within the U.S. This was helpful when creating seasonal buckets in our analysis, differentiating flight congestion depending on the time of year (season). We also differentiated Pre-Covid, Covid, and Post-Covid to see if there were any differences in the three time periods.

# I. Distribution of Flight Delays and Cancellations Across Different Airlines, Airports, and Routes

```{r, echo=FALSE}
n_unique_airlines <- flights %>%
summarise(num_airlines = n_distinct(op_unique_carrier))
```

## Visualisations of Airlines by Number of Flights 
```{r, echo=FALSE}
airline_counts <- flights %>%
count(op_unique_carrier)

ggplot(airline_counts,
aes(x = reorder(op_unique_carrier, n), y = n)) +
geom_col(fill = "darkgreen") +
coord_flip() +
labs(title = "Number of Flights by Airline (2024)",
x = "Airline", y = "Flight Count") +
theme_minimal()
```

In the graph above, we can see that Southwest Airlines (WN) had the most number of flights in our sample dataset, with a little over 2000, followed by Delta Airlines (DL) and American Airlines (AA) with approximately 1400 flights. 

## Total Miles Travelled by Each Airline
```{r, echo=FALSE}
miles_by_airline <- flights %>%
group_by(op_unique_carrier) %>%
summarise(total_miles = sum(distance, na.rm = TRUE)) %>%
arrange(desc(total_miles))

ggplot(miles_by_airline,
aes(x = reorder(op_unique_carrier, total_miles),
y = total_miles)) +
geom_col(fill = "darkgreen") +
coord_flip() +
labs(title = "Total Miles Traveled by Each Airline (2024)",
x = "Airline",
y = "Total Miles") +
theme_minimal()
```

Next, we see that this graph, which ranks the airlines based on their sum total of miles travelled, corroborates what we see in the first graph. Southwest Airlines is at the top with the most amount of miles travelled at more than 1.5 million, while Delta and American Airlines approximately travelled 1.4 million miles. 

## Average Arrival Delay by Airport 
```{r, include=FALSE}
airport_delay <- flights %>%
group_by(origin) %>%
summarise(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
arrange(desc(avg_arr_delay))
```

Moving onto arrival delay analysis, the table above shows that average arrival delays (in minutes) vary widely across airports, with a small number of locations experiencing exceptionally high delays. Victoria Regional Airport (VCT) recorded the highest average arrival delay at 152 minutes, followed by Flagstaff Pulliam Airport (FLG) at 117 minutes, and several others exceeding 100 minutes. While most airports fall below these extremes, these particular airports are significantly higher than the national average of 66 minutes, which suggests that weather patterns or infrastructure limitations can significantly affect performance. Overall, the distribution of delays indicates that chronic delays are concentrated in a subset of airports rather than evenly spread across the national aviation network.

## Average Departure Delay by Airport 
```{r, include=FALSE}
airport_dep_delay <- flights %>%
group_by(origin) %>%
summarise(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
arrange(desc(avg_dep_delay))
```

The examination of average departure delays across U.S. airports reveals a similar pattern to arrival delays, with a small group of airports experiencing disproportionately high disruptions. Victoria Regional Airport (VCT) again ranks highest, with an average departure delay of 154 minutes, followed by Kodiak Benny Benson State Aiport (116 minutes) and Minot International Aiport (112 minutes). Several others report averages exceeding 100 minutes, highlighting persistent congestion or operational challenges at these locations. While the majority of airports exhibit lower delay levels, these results suggest that there may be some structural or environmental issues that may uniquely impact specific airports. 


## Arrival and Departure Delay by Airline 
```{r, echo=FALSE}
# 1. Prepare data (Aggregate by Airline Carrier Code)
airline_delay_combined <- flights %>%
  # Group by the correct carrier code column
  group_by(op_unique_carrier) %>%
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    # Calculate overall average for consistent sorting
    overall_avg_delay = mean(c(arr_delay, dep_delay), na.rm = TRUE)
  ) %>%
  # 2. Pivot to long format
  pivot_longer(
    cols = c(avg_arr_delay, avg_dep_delay),
    names_to = "delay_type",
    values_to = "avg_delay"
  ) %>%
  # 3. Rename delay types for the legend
  mutate(
    delay_type = recode(delay_type,
                        "avg_arr_delay" = "Arrival",
                        "avg_dep_delay" = "Departure")
  )

# 4. Plot with Stacked Bars and Patterns
ggplot(airline_delay_combined,
       # Use op_unique_carrier for reordering and as the y-axis label
       aes(x = reorder(op_unique_carrier, overall_avg_delay),
           y = avg_delay, 
           fill = delay_type, 
           pattern = delay_type)) +
  
  # Use position = "stack" for additive bars
  geom_col_pattern(position = "stack", 
                   color = "black", # Outline color
                   # Pattern settings
                   pattern_fill = "black",
                   pattern_angle = 45,
                   pattern_density = 0.05,
                   pattern_spacing = 0.02) +
  
  # Manual scales to define colors and patterns
  scale_fill_manual(values = c("Departure" = "steelblue", 
                               "Arrival" = "white")) +
  
  scale_pattern_manual(values = c("Departure" = "none",
                                  "Arrival" = "stripe")) +
  
  coord_flip() + # Make it a horizontal bar chart
  
  labs(title = "Average Delay by Airline Carrier",
       x = "Airline Carrier Code", # Label now shows the carrier code
       y = "Average Delay (minutes)",
       fill = "Delay Type") + guides(pattern = "none") + theme_minimal()
```

The bar chart illustrates the average delay times for major airline carriers, segmented by departure (solid blue) and arrival (hatched) delays. There is significant variance in operational performance across the industry: American Airlines (AA) exhibits the highest cumulative delay, approaching 40 minutes, followed closely by Frontier (F9) and PSA Airlines (OH). In contrast, Republic Airways (YX) stands out as a distinct anomaly; despite recording minor positive departure delays, it shows negative arrival delays, indicating a consistent pattern of arriving ahead of schedule. For the majority of other carriers, however, departure delays serve as the dominant component, compounding into substantial arrival delays.

## Route Level Delay Analysis 
```{r, echo=FALSE}
route_delay_prob <- flights %>%
  group_by(origin, dest) %>%
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    flights = n(),
    .groups = 'drop' # Recommended to manage groups
  ) %>%
  filter(flights >= 10) %>%
  pivot_longer(
    cols = c(avg_arr_delay, avg_dep_delay),
    names_to = "delay_type",
    values_to = "avg_delay"
  ) %>%
  mutate(
    delay_type = recode(delay_type,
                        "avg_arr_delay" = "Arrival Delay",
                        "avg_dep_delay" = "Departure Delay"),
    route = paste(origin, "→", dest)
  )

# --- Top Routes Calculation (FIXED: Added na.rm = TRUE) ---
# Calculate the maximum delay (either arrival or departure) for each route
top_routes_data <- route_delay_prob %>%
  group_by(route) %>%
  summarise(
    # FIX: Add na.rm = TRUE to suppress the warning if any avg_delay is missing
    max_delay = max(avg_delay, na.rm = TRUE), 
    .groups = 'drop'
  ) %>%
  # Select the top 15 routes based on that max delay
  slice_max(max_delay, n = 15) 

# FIX: Join the max_delay column back to the main data for consistent sorting
plot_data <- route_delay_prob %>%
  # Filter the main data to include only the top 20 routes
  inner_join(top_routes_data, by = "route")

# --- Plotting --- (Rest of the plot code remains the same)
ggplot(plot_data, 
       aes(x = reorder(route, max_delay), 
           y = avg_delay, 
           fill = delay_type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Top 15 Most Delayed Routes (Arrival & Departure, 2024)",
    x = "Route", 
    y = "Average Delay (minutes)", 
    fill = "Delay Type"
  ) +
  scale_fill_manual(
    values = c("Arrival Delay" = "skyblue",
               "Departure Delay" = "#FBC02D")
  ) +
  theme_minimal()
```

This bar chart details the top 15 most delayed flight routes in 2024. The route from Newark (EWR) to Chicago (ORD) suffers the most severe disruptions with average departure delays approaching 90 minutes and arrival delays exceeding 75 minutes. A consistent trend across the data is that departure delays (yellow) generally outpace arrival delays (blue), indicating that while aircraft often recover some time mid-flight, initial ground setbacks remain the primary driver of lateness. While major hubs like Atlanta (ATL) and Dallas-Fort Worth (DFW) feature prominently in the top rankings, the severity of delays tapers off significantly after the top few routes, with the bottom half of the list averaging between 25 and 35 minutes of delay.

## Route Level Cancellation Analysis
```{r, echo=FALSE}
# --- Data Preparation ---
route_cancel_prob <- flights %>%
  group_by(origin, dest) %>%
  summarise(
    cancel_rate = mean(cancelled == 1, na.rm = TRUE),
    flights = n(),
    .groups = 'drop'
  ) %>%
  filter(flights >= 10) %>% 
  mutate(route = paste(origin, "→", dest)) %>%
  arrange(desc(cancel_rate))

# --- Top 15 Routes Calculation ---
top_cancelled_routes <- route_cancel_prob %>%
  slice_max(order_by = cancel_rate, n = 10, with_ties = FALSE)  # <-- only top 15

# --- Plotting (Fixed) ---
ggplot(top_cancelled_routes,
       aes(x = reorder(route, cancel_rate), y = cancel_rate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Top 10 Most Cancelled Routes (2024)", 
    x = "Route", 
    y = "Probability of Cancellation"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.margin = unit(c(1, 1, 1, 3), "lines"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

This bar chart details the top 10 most cancelled routes of 2024. The route with the highest probability of cancellation is the Tampa (TPA) to Atlanta (ATL) route, which faces a cancellation probability of nearly 17%. This rate is significantly higher than the rest of the field, nearly doubling the roughly 10% cancellation rates seen on the next most affected routes, Newark (EWR) to Fort Lauderdale (FLL) and Boston (BOS) to Orlando (MCO). It is also interesting that while TPA to ATL is the most cancelled route here, it did not appear in the most delayed routes graph, while the reverse route of ATL to TPA was present. This suggests that when operational issues hit this specific route, airlines are more likely to cancel this flight entirely rather than delay it. This could be due to the fact that when bad weather or congestion hits a major hub like Atlanta, operational teams often sacrifice these shorter regional flights first; because there are multiple daily departures, it is far easier to rebook passengers on the next plane than it would be to cancel a long-haul flight, essentially making the TPA-ATL route a strategic casualty to keep the broader system moving. 

# II. Temporal Patterns in Flight Cancellations and Delays 
```{r}
# ---- Helper function: extract hour from HHMM format ----
extract_hour <- function(x) {
  x_num <- suppressWarnings(as.numeric(x))
  ifelse(
    is.na(x_num),
    NA_integer_,
    floor(x_num / 100) %% 24L
  )
}

# ---- Feature Engineering on flights dataset ----
flights_temp <- flights %>%
  mutate(
    # BTS data: day_of_week = 1 (Mon) ... 7 (Sun)
    day_of_week = factor(
      day_of_week,
      levels = 1:7,
      labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    ),

    # numeric formats
    dep_delay = as.numeric(dep_delay),
    arr_delay = as.numeric(arr_delay),
    cancelled = as.integer(cancelled),

    # scheduled departure hour (0–23)
    dep_hour = extract_hour(crs_dep_time),

    # bucket hours into coarse time-of-day groups
    dep_period = case_when(
      dep_hour >= 5  & dep_hour < 10 ~ "Early AM (5–9)",
      dep_hour >= 10 & dep_hour < 14 ~ "Late AM (10–13)",
      dep_hour >= 14 & dep_hour < 18 ~ "Afternoon (14–17)",
      dep_hour >= 18 & dep_hour < 22 ~ "Evening (18–21)",
      TRUE                           ~ "Night (22–4)"
    ),

    # ordered factor for period visualization
    dep_period = factor(
      dep_period,
      levels = c(
        "Night (22–4)",
        "Early AM (5–9)",
        "Late AM (10–13)",
        "Afternoon (14–17)",
        "Evening (18–21)"
      )
    )
  )
```


## Patterns by Day of Week 
```{r, include=FALSE}
dow_summary <- flights_temp %>%
  group_by(day_of_week) %>%
  summarise(
    n_flights      = n(),
    mean_dep_delay = mean(dep_delay, na.rm = TRUE),
    mean_arr_delay = mean(arr_delay, na.rm = TRUE),
    cancel_rate    = mean(cancelled, na.rm = TRUE),
    .groups = "drop"
  )

dow_summary
```

```{r}
#plot: average departure delay by day of week
ggplot(dow_summary, aes(x = day_of_week, y = mean_dep_delay)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Departure Delay by Day of Week",
    x = "Day of Week",
    y = "Mean Departure Delay (minutes)"
  )

#plot: cancellation rate by day of week
ggplot(dow_summary, aes(x = day_of_week, y = cancel_rate)) +
  geom_col(fill = "orange") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Cancellation Rate by Day of Week",
    x = "Day of Week",
    y = "Cancellation Rate"
  )
```

Next, we move onto temporal pattern analysis for delays and cancellations. This bar chart of average departure trends highlights a midweek sweet spot for travelers, with Tuesday and Wednesday showing the lowest average delays at roughly 10 minutes. In contrast, schedule reliability worsens toward the weekend, peaking on Saturday with an average delay exceeding 16 minutes. This pattern likely reflects the cumulative impact of network congestion: midweek flights benefit from lighter business travel, whereas weekend surges of leisure travelers, combined with delays accumulated earlier in the week, create a compounding effect. Notably, although many travelers assume Friday or Sunday are the busiest days, the data indicate that Saturday experiences the longest delays, emphasizing that peak passenger volume does not always coincide with peak operational disruption.

This bar chart examining 2024 cancellation rates by day of the week reveals a surprising paradox for travelers. While Wednesday and Thursday generally offer better on-time performance in terms of delays, they are ironically the riskiest days for outright cancellations, with rates peaking near 1.45%. In contrast, Sunday, often assumed to be chaotic due to high passenger volume, has the lowest cancellation probability, below 0.9%. This pattern suggests that airlines may strategically cancel or consolidate flights on lower-demand midweek days to optimize operational efficiency, while making extra effort to maintain schedules on high-demand weekends to avoid stranding large numbers of passengers.

## Patterns by Hour of Day
```{r}
hour_summary <- flights_temp %>%
  group_by(dep_hour) %>%
  summarise(
    n_flights      = n(),
    mean_dep_delay = mean(dep_delay, na.rm = TRUE),
    mean_arr_delay = mean(arr_delay, na.rm = TRUE),
    cancel_rate    = mean(cancelled, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(dep_hour))

#plot: average departure delay by departure hour
ggplot(hour_summary, aes(x = dep_hour, y = mean_dep_delay)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Average Departure Delay by Hour of Day",
    x = "Scheduled Departure Hour",
    y = "Mean Departure Delay (minutes)"
  )

#plot: cancellation rate by departure hour
ggplot(hour_summary, aes(x = dep_hour, y = cancel_rate)) +
  geom_line(color = "orange", linewidth = 1) +
  geom_point(color = "orange", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Cancellation Rate by Hour of Day",
    x = "Scheduled Departure Hour",
    y = "Cancellation Rate"
  )
```

The first line graph shows average departure delays by hour. Most flights have the shortest delays between 5:00 AM and 11:00 AM, staying under 10 minutes on average. Delays gradually increase during the day, peaking at over 25 minutes around 8:00 PM. An exception is 4:00 AM, where delays jump to over 50 minutes. This spike likely happens because there are very few flights at that hour, so one delay, such as a late crew or maintenance issue, can make the average very high.

This second line graph shows cancellation rates by hour. While there are almost no cancellations between midnight and 4:00 AM, rates jump at 5:00 AM to over 2%, likely due to overnight maintenance issues or late-arriving planes. The highest risk happens late at night, around 10:00 PM, when cancellations exceed 3%. These late-night cancellations usually occur because delays have built up during the day, crews have reached their maximum work hours, or airports close for the night, making the last flights of the day the riskiest for travelers.


## Patterns by Time-of-Day Buckets
```{r}
period_summary <- flights_temp %>%
  group_by(dep_period) %>%
  summarise(
    n_flights      = n(),
    mean_dep_delay = mean(dep_delay, na.rm = TRUE),
    mean_arr_delay = mean(arr_delay, na.rm = TRUE),
    cancel_rate    = mean(cancelled, na.rm = TRUE),
    .groups = "drop"
  )

#plot: average departure delay by time-of-day bucket
ggplot(period_summary, aes(x = dep_period, y = mean_dep_delay)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Departure Delay by Time of Day",
    x = "Time of Day",
    y = "Mean Departure Delay (minutes)"
  ) +
  coord_flip()  # flip for readability

#plot: cancellation rate by time-of-day bucket
ggplot(period_summary, aes(x = dep_period, y = cancel_rate)) +
  geom_col(fill = "orange") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Cancellation Rate by Time of Day",
    x = "Time of Day",
    y = "Cancellation Rate"
  ) +
  coord_flip()
```

The first bar chart illustrating average departure delay by time of day confirms that evening time (18-21) is the most disrupted travel window, suffering the highest delays at nearly 20 minutes. In contrast, the AM (5–9) block offers the most reliable performance, with average delays significantly lower at roughly 7.5 minutes. While delays climb steadily through the Late AM (10–13) and Afternoon (14–17) blocks, an interesting anomaly appears in the Night (22–4) category; despite low passenger volumes, delays here remain stubbornly high at around 17 minutes, supporting the results seen in other graphs about night time delays. 

The second bar chart categorizing cancellation rates by time of day identifies a critical block for travelers: the Night (22–4) block. Flights scheduled during these late hours face a cancellation rate approaching 2.8%, nearly double the rate of any other time block. While the Late AM (10–13) block appears to be the safest bet with cancellation rates dipping below 1.0%, the risk climbs again in the Afternoon (14–17) to roughly 1.4%. The data strongly suggests that flying very late is a gamble; unlike daytime cancellations where rebooking options exist, a cancellation in the 22:00–04:00 window often means being stranded overnight as airline operations wind down for the day.

# III. Geographical Patterns in Flight Cancellations and Delays
```{r, include=FALSE}
airports_us <- read_csv("airports.csv")
```

```{r}
airports_needed <- flights %>%
  select(origin, dest) %>%
  pivot_longer(cols = c(origin, dest), names_to = "type", values_to = "IATA") %>%
  distinct(IATA)

airports_2024 <- airports_us %>%
  filter(IATA %in% airports_needed$IATA)

flights_geo <- flights %>%
  left_join(airports_2024 %>% select(IATA, LATITUDE, LONGITUDE), by = c("origin" = "IATA")) %>%
  rename(origin_lat = LATITUDE, origin_lon = LONGITUDE) %>%
  left_join(airports_2024 %>% select(IATA, LATITUDE, LONGITUDE), by = c("dest" = "IATA")) %>%
  rename(dest_lat = LATITUDE, dest_lon = LONGITUDE)
```

```{r}
# Cleaning Dataset
orig_df <- flights_geo %>%
  transmute(
    airport = origin,
    city = origin_city_name,
    state = origin_state_nm,
    lat = origin_lat,
    lon = origin_lon,
    cancelled = cancelled,
    diverted = diverted,
    arr_delay = arr_delay,
    dep_delay = dep_delay
  )

dest_df <- flights_geo %>%
  transmute(
    airport = dest,
    city = dest_city_name,
    state = dest_state_nm,
    lat = dest_lat,
    lon = dest_lon,
    cancelled = cancelled,
    diverted = diverted,
    arr_delay = arr_delay,
    dep_delay = dep_delay
  )

air_all <- bind_rows(orig_df, dest_df)

air_all_clean <- air_all %>%
  filter(!is.na(lat) & !is.na(lon))

airport_stats <- air_all_clean %>%
  group_by(airport, city, state, lat, lon) %>%
  summarise(
    total_flights = n(),
    cancelled_count = sum(cancelled == 1, na.rm = TRUE),
    cancelled_rate = cancelled_count / total_flights,
    avg_arr_delay = mean(arr_delay[cancelled == 0 & diverted == 0], na.rm = TRUE)
  ) %>%
  ungroup()

airport_stats <- airport_stats %>%
  mutate(
    avg_arr_delay = ifelse(is.nan(avg_arr_delay), NA, avg_arr_delay)
  )

# cleaner visualization by only keeping airports with 20+ flights
airport_stats_plot <- airport_stats %>%
  filter(total_flights >= 20)

us_states <- map_data("state")

# Cancellation rate map
map_cancel <- ggplot() +
  geom_polygon(data = us_states,
               aes(x = long, y = lat, group = group),
               fill = "gray95", color = "gray80") +
  geom_point(data = airport_stats_plot,
             aes(x = lon, y = lat, size = total_flights, color = cancelled_rate),
             alpha = 0.8) +
  scale_color_viridis_c(option = "plasma",
                        labels = percent_format(accuracy = 0.1),
                        name = "Cancellation Rate") +
  scale_size(range = c(1, 6), name = "Total Flights") +
  coord_fixed(1.3) +
  labs(title = "U.S. Airport Cancellation Rates (2024)") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

# Average arrival delay map
map_delay <- ggplot() +
  geom_polygon(data = us_states,
               aes(x = long, y = lat, group = group),
               fill = "gray95", color = "gray80") +
  geom_point(data = airport_stats_plot,
             aes(x = lon, y = lat, size = total_flights, color = avg_arr_delay),
             alpha = 0.8) +
  scale_color_viridis_c(option = "magma",
                        name = "Avg Arrival Delay (min)",
                        direction = -1) +
  scale_size(range = c(1, 6), name = "Total Flights") +
  coord_fixed(1.3) +
  labs(title = "U.S. Airport Average Arrival Delay (2024)") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank())

options(repr.plot.width = 14, repr.plot.height = 8)
map_cancel
map_delay

```

We now move onto geographic patterns of flights cancellation & analysis. This geographic heatmap of cancellation rates reveals that flight reliability is highly dependent on location, with risk concentrated in specific regional pockets rather than evenly distributed across the country. The bright yellow hotspots indicating cancellation rates approaching 9% are visible in areas like the Northeast and parts of the Southeast, contrasting sharply with the stable dark purple markers (near 0% cancellations) that dominate the West Coast. Interestingly, the largest circles representing major hubs like Atlanta and Chicago are generally not the brightest yellow, suggesting that while they handle the most volume, it is often the smaller or secondary airports that suffer the highest probability of having flights scrapped entirely, likely serving as the first cuts airlines make during operational stress.

The second map shows average arrival delays across the U.S. and highlights a clear East vs. West pattern. Airports in the Eastern U.S., particularly in the Northeast corridor and Florida, experience the longest delays, often 15–25 minutes on average. This is likely due to higher traffic density, congested airspace, and frequent weather disruptions in the region. In contrast, airports in the Western U.S. mostly see shorter delays of 0–10 minutes, reflecting lower traffic volumes and less congested routes. The map also shows that major hub airports, such as those in New York, Atlanta, and Miami, tend to have darker markers, confirming that large, busy airports contribute disproportionately to national delays. Overall, the visualization suggests that both geographic location and airport size strongly influence arrival punctuality, and that travelers flying west of the Mississippi generally have a higher chance of arriving on time.

```{r}
top_cancel <- airport_stats %>%
  arrange(desc(cancelled_count)) %>%
  slice_head(n = 15)   # top 15 airports

ggplot(top_cancel, aes(x = reorder(airport, cancelled_count), 
                       y = cancelled_count)) +
  geom_col(fill = viridis::viridis(15)) +
  coord_flip() +
  labs(
    title = "Top Airports by Number of Cancellations (2024)",
    x = "Airport",
    y = "Total Cancellations"
  ) +
  theme_minimal(base_size = 12)
```

This bar chart ranks airports by the total number of cancellations, showing Dallas-Fort Worth (DFW) at the top with over 20 cancellations in the sample dataset. The runner-up, Newark (EWR), has about 14, while LaGuardia (LGA) and Atlanta (ATL) follow with around 10 each. DFW’s cancellations are much higher than other major hubs like Chicago (ORD) or Detroit (DTW), which have roughly 5 each, highlighting that cancellations are heavily concentrated at DFW. Several factors could contribute to this, including DFW’s high flight volume, its role as a major connecting hub where delays span across the network, weather disruptions in the region, and operational constraints such as runway capacity or staffing shortages. It is also interesting to note that while TPA -> ATL had the highest cancellation rate on a previous chart, DFW has the highest total number of cancellations. This suggests that TPA’s issue is likely route-specific, whereas DFW faces broader operational challenges affecting many flights.

```{r}
airport_delay_counts <- air_all_clean %>%
  mutate(is_delayed = arr_delay > 15) %>%
  group_by(airport) %>%
  summarise(delayed_flights = sum(is_delayed, na.rm = TRUE)) %>%
  inner_join(airport_stats, by = "airport") %>%
  arrange(desc(delayed_flights)) %>%
  slice_head(n = 15)

ggplot(airport_delay_counts, 
       aes(x = reorder(airport, delayed_flights),
           y = delayed_flights)) +
  geom_col(fill = viridis::viridis(15, option = "inferno")) +
  coord_flip() +
  labs(
    title = "Airports with the Most Delayed Flights (>15 min)",
    x = "Airport",
    y = "Number of Delayed Flights"
  ) +
  theme_minimal(base_size = 12)
```

This bar chart shows airports ranked by the number of flights delayed more than 15 minutes, a common threshold to highlight delays that really affect passengers. Dallas-Fort Worth (DFW) has the most delays, with over 225 flights, followed by Atlanta (ATL) and Denver (DEN) with about 180–190 each. Big central hubs like Chicago (ORD) and Charlotte (CLT) also have many delays because they handle a lot of connecting flights, tight schedules, and delays that spread from earlier flights. Other factors, like bad weather, crowded airspace, and limited runway or gate space, make delays worse. In contrast, busy coastal airports like JFK and Boston (BOS) have fewer major delays, showing that the biggest problems happen at central hubs with complex flight networks.

```{r}
#select airports of interest 
airports_of_interest <- c("DFW", "LGA", "EWR", "LAX")

#filter flights and gather delay reasons 
delay_long <- flights_geo %>%
  filter(origin %in% airports_of_interest | dest %in% airports_of_interest) %>%
  mutate(airport = if_else(origin %in% airports_of_interest, origin, dest)) %>%
  select(
    airport,
    carrier_delay,
    weather_delay,
    nas_delay,
    security_delay,
    late_aircraft_delay
  ) %>%
  pivot_longer(
    cols = c(carrier_delay, weather_delay, nas_delay, security_delay, late_aircraft_delay),
    names_to = "delay_type",
    values_to = "delay_minutes"
  ) %>%
  mutate(
    delay_minutes = replace_na(delay_minutes, 0)
  ) %>%
  group_by(airport, delay_type) %>%
  summarise(total_delay = sum(delay_minutes), .groups = "drop")

#convert delay type labels to nicer names 
delay_long$delay_type <- recode(delay_long$delay_type,
                                carrier_delay = "Carrier",
                                weather_delay = "Weather",
                                nas_delay = "NAS",
                                security_delay = "Security",
                                late_aircraft_delay = "Late Aircraft")

#pie charts for each airport 
ggplot(delay_long, aes(x = "", y = total_delay, fill = delay_type)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~ airport) +
  scale_fill_viridis_d(name = "Delay Reason") +
  labs(
    title = "Delay Reasons for Selected Airports (DFW, LGA, EWR, LAX)",
    y = "Total Delay Minutes",
    x = NULL
  ) +
  theme_void() +
  theme(
    strip.text = element_text(size = 14),
    plot.title = element_text(size = 16, hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(10, 5, 10, 10)
  )

```

This set of pie charts breaks down delay causes for four major hubs, revealing a clear distinction between network problems and local problems. We chose to focus on the top three airports with the most delays in our dataset: Atlanta (ATL), Denver (DEN), and Dallas-Fort Worth (DFW); as well as Los Angeles (LAX) because we live in Claremont and often commute through LAX. The three busiest connecting hubs share a similar profile dominated by late aircraft, indicating that their main reasons for delays come from incoming flights arriving behind schedule. In contrast, LAX stands apart with a dominant carrier delay share, suggesting that disruptions there are more often caused by airline-specific issues such as maintenance or crew logistics rather than system-wide congestion. Additionally, NAS (National Aviation System) delays, which include heavy air traffic volume and minor weather impacts, play a noticeable role at DEN and DFW, while weather delays remain visible at the interior hubs compared to the generally consistent weather at LAX. 

# IV. Reasons for Delays and Cancellations
```{r, echo=FALSE}
delay_reasons_long <- flights %>%
  select(carrier_delay, weather_delay, nas_delay, security_delay, late_aircraft_delay) %>%
  pivot_longer(cols = everything(),
               names_to = "delay_type",
               values_to = "minutes") %>%
  filter(minutes > 0)

delay_summary <- delay_reasons_long %>%
  group_by(delay_type) %>%
  summarise(total_delay = sum(minutes, na.rm = TRUE)) %>%
  arrange(desc(total_delay))

delay_summary <- delay_summary %>%
  mutate(prop = total_delay / sum(total_delay))

ggplot(delay_summary, aes(x = "", y = prop, fill = delay_type)) +
  geom_col(width = 0.5) +
  coord_polar(theta = "y") +
  labs(title = "Proportion of Total Delay by Reason (2024)", fill = "Delay Type") +
  scale_y_continuous(labels = scales::percent) +
  theme_void()

delay_summary_avg <- delay_reasons_long %>%
  group_by(delay_type) %>%
  summarise(avg_delay = mean(minutes, na.rm = TRUE))
```

This pie chart breaking down the main causes of flight delays in 2024 shows that the biggest factor is not weather, but the ripple effect of earlier delays. Late aircrafts makes up the largest share of total delay minutes, showing that when a plane is late, it causes delays for all subsequent flights using that aircraft. The second-largest factor is carrier delay, which includes airline-controlled issues like maintenance, cleaning, or crew scheduling. Together, these two operational causes have a much bigger impact than weather delays or National Air System Delays, challenging the common idea that storms are the main reason for getting stuck on the tarmac. This pattern aligns with what we saw in the hub-specific pie charts, where late aircraft delays dominated at major connecting airports like ATL, DEN, and DFW, while carrier-related issues were particularly prominent at LAX.

Overall, our exploratory analysis reveals clear patterns in flight delays and cancellations across U.S. airlines, airports, and routes. Delays are concentrated at major connecting hubs, particularly DFW, ATL, and DEN, with interior airports experiencing higher disruption than most coastal airports. Operational factors like late aircraft and carrier delays consistently outweigh weather and National Air System delays, emphasizing the cascading nature of flight schedules. Temporal patterns show that midweek and early morning flights are generally more reliable, while weekends and late-night flights carry higher risks of delay or cancellation. Geographically, the Eastern U.S. and heavily trafficked hubs see the most significant delays, while Western airports benefit from lighter congestion. These insights collectively paint a detailed picture of when, where, and why flights experience operational difficulties.


# V. Flight Delay Predictions

Building on these insights, the next step of our project focuses on predicting flight delays using supervised learning. By leveraging the patterns identified in the exploratory analysis, such as airline, airport, route, day of the week, and time of day, we aim to construct a model capable of estimating the likelihood of a flight being delayed. This predictive approach will allow us to move beyond descriptive insights, providing actionable information for travelers and airlines to anticipate and mitigate disruptions. 

## Feature Engineering: Day of Week × Time of Day
We believe that flight delays are not determined by the day of the week or the time of day independently, but rather by the interaction between the two. For example, a Monday morning departure may face very different operational conditions than a Sunday morning departure, and weekday afternoon flights are often subject to peak congestion that does not occur at the same time on weekends. 

By constructing an interaction term between day of week and time of day, we want to capture these compound scheduling effects that would otherwise be missed by treating the variables separately. This interaction allows our model to distinguish between structurally different flight environments, such as weekday peak-hour traffic versus off-peak weekend operations, thereby providing a more realistic representation of temporal congestion patterns in air travel.

```{r}
FE <- flights %>%
  mutate(
    #ensure numeric delay variables
    dep_delay = as.numeric(dep_delay),
    arr_delay = as.numeric(arr_delay),
    cancelled = as.integer(cancelled),

    #convert day_of_week (1–7) to labels
    day_of_week = factor(
      as.numeric(day_of_week),
      levels = 1:7,
      labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    ),

    #extract scheduled departure hour
    dep_hour = extract_hour(crs_dep_time),

    #create time-of-day buckets
    time_of_day = case_when(
      dep_hour >= 5  & dep_hour < 10 ~ "Early AM",
      dep_hour >= 10 & dep_hour < 14 ~ "Late AM",
      dep_hour >= 14 & dep_hour < 18 ~ "Afternoon",
      dep_hour >= 18 & dep_hour < 22 ~ "Evening",
      TRUE                           ~ "Night"
    ),

    #ordered factor for time_of_day
    time_of_day = factor(
      time_of_day,
      levels = c("Night", "Early AM", "Late AM", "Afternoon", "Evening")
    ),

    #interaction feature: Day of Week × Time of Day
    dow_period = interaction(
      day_of_week,
      time_of_day,
      sep = " x ",
      drop = TRUE
    )
  )
```


```{r, include=FALSE}
AirPassengers <- read_csv("/Users/chrysanthemum/Final Project/AirPassengers.csv")
```


## Feature Engineering: Travel Season

We hypothesize that domestic air passenger demand exhibits a strong seasonal pattern driven by travel behavior such as holidays, school schedules, and weather conditions. To explore this, we analyze monthly domestic passenger data from the U.S. Bureau of Transportation Statistics, aggregated across 2015-2025. 

The data reveal a clear and recurring pattern: passenger volumes are lowest in January and February, rise sharply beginning in March, and remain elevated throughout the spring and summer months, peaking between March and August. Demand then declines during the fall and early winter period from September to December, though it remains higher than the early-year trough. 

```{r}
# Ensure numeric
AirPassengers <- AirPassengers %>%
  mutate(
    Year = as.numeric(as.character(Year)),
    Month = as.numeric(as.character(Month)),
    DOMESTIC = as.numeric(DOMESTIC)
  )

# Now filter — only works if years exist
df <- AirPassengers %>%
  filter(Year >= 2015, Year <= 2025)

if(nrow(df) == 0){
  stop("❌ No data exists between 2015 and 2025 in AirPassengers.")
}

# Define periods
df <- df %>%
  mutate(
    Period = case_when(
      Year >= 2015 & Year <= 2019 ~ "2015–2019",
      Year >= 2020 & Year <= 2022 ~ "2020–2022",
      Year >= 2023 & Year <= 2025 ~ "2023–2025"
    )
  )

# Compute averages
monthly_period_avg <- df %>%
  group_by(Period, Month) %>%
  summarize(AvgDomestic = mean(DOMESTIC, na.rm = TRUE), .groups = "drop")

ggplot(monthly_period_avg, aes(x = Month, y = AvgDomestic, color = Period)) +
  geom_line(size = 1.4) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:12) +
  coord_cartesian(ylim = c(2.5e7, 1e8)) +
  labs(
    title = "Domestic Passengers by Month: 3 Time Period Averages",
    x = "Month",
    y = "Average Domestic Passengers (thousands)",
    color = "Period"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title   = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text.x  = element_text(size = 9),
    axis.text.y  = element_text(size = 9),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9)
  )
```

```{r}
#feature engineering - travel season 
flights_temp_season <- flights_temp %>%
  mutate(
    Month_num = as.integer(as.character(month)),  
    Season = case_when(
      Month_num %in% 3:8  ~ "peak",        
      Month_num %in% 9:12 ~ "moderate",   
      Month_num %in% 1:2  ~ "shoulder",    
      TRUE ~ NA_character_
    ),
    Season = factor(Season, levels = c("shoulder", "moderate", "peak"))
  )
```


Based on this empirical pattern, we construct a categorical Season variable by grouping months into three demand regimes: shoulder season (January–February), peak season (March–August), and moderate season (September–December). 

## Use of Boosting for Delay Prediction and Resulting Performance
We apply a gradient boosting model to predict whether a flight experiences a departure delay of 15 minutes or more using airline, day of week, departure hour, travel season, and the interaction between day of week and time of day (dow_period) as predictors. Boosting is well suited for this task because it can capture nonlinear effects and complex temporal interactions that simpler models may overlook.

In practice, while the boosting model achieves high overall accuracy, this performance is driven largely by the predominance of on-time flights in the dataset. The model demonstrates strong sensitivity to on-time flights but very limited ability to identify delayed flights, reflecting the underlying class imbalance and the limited predictive signal in the available scheduling variables. Adjusting probability thresholds improves detection of delayed flights but highlights the inherent difficulty of the prediction task rather than a deficiency in the modeling approach.

```{r}
library(gbm)

#keep only flights that were not cancelled
flights_nocancel <- flights_temp_season %>%
  filter(cancelled == 0)

#prepare dataset 
flights_boosting <- flights_nocancel %>%
  mutate(
    dep_delay   = as.numeric(dep_delay),
    delayed15   = if_else(dep_delay >= 15, "delayed", "on_time"),
    delayed15   = factor(delayed15, levels = c("on_time", "delayed")),
    airline     = factor(op_unique_carrier),
    day_of_week = factor(day_of_week),
    dep_hour    = as.numeric(dep_hour),
    dow_period = interaction(day_of_week, dep_period, sep = " x ", drop = TRUE),
    dow_period = factor(stringr::str_squish(as.character(dow_period))),
    season = factor(Season, levels = c("shoulder", "moderate", "peak")) 

  ) %>%
  drop_na(delayed15, airline, day_of_week, dep_hour, dow_period, season)

```

```{r}
#train/test split
set.seed(123)
train_index <- createDataPartition(flights_boosting$delayed15, p = 0.8, list = FALSE)

train_data <- flights_boosting[train_index, ]
test_data  <- flights_boosting[-train_index, ]

#align factor levels between train and test
for (v in c("airline", "day_of_week", "dow_period", "season")) {
  test_data[[v]] <- factor(test_data[[v]], levels = levels(train_data[[v]]))
}

test_data <- test_data %>%
  drop_na(airline, day_of_week, dow_period, season)

```

```{r, include=FALSE}
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

set.seed(123)
gbm_model <- train(
  delayed15 ~ airline + day_of_week + dep_hour + dow_period + season,
  data      = train_data,
  method    = "gbm",
  trControl = ctrl,
  metric    = "ROC",
  verbose   = FALSE
)

gbm_model
```

```{r, include=FALSE}
#predicted probabilities
gbm_prob <- predict(gbm_model, newdata = test_data, type = "prob")

#evaluate with confusion matrix 
gbm_pred_35 <- if_else(
  gbm_prob$delayed >= 0.3,
  "delayed",
  "on_time"
) %>% factor(levels = c("on_time", "delayed"))

confusionMatrix(gbm_pred_35, test_data$delayed15)

```

```{r}
library(pROC)

roc_obj <- roc(
  response  = test_data$delayed15,
  predictor = gbm_prob$delayed,
  levels    = c("on_time", "delayed"),
  direction = "<"
)

auc_val <- auc(roc_obj)

# increase top margin 
par(mar = c(5, 4, 6, 2)) # bottom, left, top, right 
plot( roc_obj, main = "", col = "steelblue", lwd = 2 ) 
title( main = 
         paste0( "ROC Curve for Boosting Model\nAUC = ", 
                      round(auc_val, 3) ), 
       line = 4 # increase this to add more vertical space 
       ) 
abline(a = 0, b = 1, lty = 2, col = "gray")

```

## Insights from Unsupervised Learning 
Given the limited discriminative performance of supervised models, we transition to unsupervised learning to better understand the structure of the feature space and diagnose the source of model limitations. Using Principal Component Analysis (PCA), we examine whether delayed flights naturally separate from on-time flights based on airline and temporal characteristics. The PCA results reveal substantial overlap between delayed and on-time flights across all principal components, indicating that these classes are not well separated in the feature space. This finding suggests that flight delays are influenced by stochastic operational factors, such as weather disruptions, aircraft availability, and air traffic control decisions, that are not captured by scheduling information alone. 

```{r}
#prepare data for PCA
pca_data <- flights_boosting %>%
  select(airline, day_of_week, dep_hour, dow_period, delayed15)

#dummy encode categorical variables
dummies <- dummyVars(delayed15 ~ ., data = pca_data)
pca_matrix <- predict(dummies, pca_data)

#run PCA (scale = TRUE is critical)
pca_res <- prcomp(pca_matrix, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  PC1 = pca_res$x[,1],
  PC2 = pca_res$x[,2],
  delayed15 = pca_data$delayed15
)

ggplot(pca_df, aes(PC1, PC2, color = delayed15)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "PCA of Flight Features",
    x = "PC1",
    y = "PC2",
    color = "Delay Status"
  ) +
  theme_minimal()
```

# VI. Predictions of Flight Delay Length
We estimate a linear regression model to predict departure delay length (in minutes) using airline fixed effects, day-of-week indicators, departure hour, and a seasonal travel indicator. Overall, the model is statistically significant (F-statistic = 8.10, p < 0.001), indicating that the included predictors jointly explain variation in departure delays. However, the model’s explanatory power is modest, with an R-squared of 0.023, suggesting that only about 2.3% of the variation in delay length is explained by these variables. This is expected given that flight delays are influenced by many unobserved operational factors such as weather, air traffic congestion, and aircraft turnaround constraints.

## Key Predictors
Departure hour is a strong and highly significant predictor of delay length. The coefficient of 1.01 implies that, holding other factors constant, each additional hour later in the day is associated with approximately one extra minute of departure delay. This result is consistent with delay propagation throughout the day, where early disruptions accumulate and affect later flights.

The seasonal indicators reveal meaningful differences in delay severity across travel seasons. Relative to the shoulder season (January–February), flights during the peak season (March–August) experience delays that are on average 6.27 minutes longer, and this effect is statistically significant (p < 0.001). In contrast, the moderate season (September–December) is associated with slightly shorter delays (–2.37 minutes), though this effect is not statistically significant. These findings align with higher demand, congestion, and capacity constraints during peak travel months.

Airline effects are present but mixed. Some carriers exhibit significantly longer delays relative to the reference airline, while others show no statistically distinguishable differences. This suggests heterogeneity in operational performance, though many airline coefficients are imprecisely estimated once time and season controls are included.

Day-of-week effects are generally small and statistically weak. While weekend flights (especially Saturdays) show slightly longer delays on average, these effects are not strongly significant, indicating that departure time and seasonality play a larger role than the day of week in explaining delay duration.

## Model Performance
Out-of-sample evaluation yields an RMSE of 42.3 minutes and a mean absolute error (MAE) of 23.0 minutes. These values indicate substantial prediction uncertainty, reflecting the inherently noisy nature of delay length outcomes. While the model captures systematic patterns related to time of day and seasonality, it is not designed to provide highly precise individual flight delay predictions.

## Summary
Overall, the results suggest that temporal and seasonal factors systematically affect the length of flight delays, with peak travel periods and later departures being particularly prone to longer delays. However, the low R-squared highlights the limitations of a linear model that excludes key operational variables. This regression should therefore be interpreted as capturing average structural effects rather than serving as a high-accuracy predictive model. 

```{r}
#prepare variables for regression: predict dep_delay (minutes)
flights_nocancel_lrm <- flights_nocancel %>%
  mutate(
    dep_delay   = as.numeric(dep_delay),      # outcome
    airline     = factor(op_unique_carrier),
    day_of_week = factor(day_of_week),
    dep_hour    = as.numeric(dep_hour),
    Season      = factor(Season, levels = c("shoulder", "moderate", "peak"))  
  ) %>%
  drop_na(dep_delay, airline, day_of_week, dep_hour, Season)                  

```

```{r}
set.seed(123)
train_index <- createDataPartition(flights_nocancel_lrm$dep_delay, p = 0.8, list = FALSE)

train_data <- flights_nocancel_lrm[train_index, ]
test_data  <- flights_nocancel_lrm[-train_index, ]

for (v in c("airline", "day_of_week", "Season")) {   
  test_data[[v]] <- factor(test_data[[v]], levels = levels(train_data[[v]]))
}

test_data <- test_data %>%
  drop_na(airline, day_of_week, Season)
```

```{r, include=FALSE}
#regression model: predict minutes of departure delay
lm_model <- lm(
  dep_delay ~ airline + day_of_week + dep_hour + Season,
  data = train_data
)

summary(lm_model)
```

```{r, include=FALSE}
delay_pred <- predict(lm_model, newdata = test_data)

rmse <- sqrt(mean((delay_pred - test_data$dep_delay)^2, na.rm = TRUE))
mae  <- mean(abs(delay_pred - test_data$dep_delay), na.rm = TRUE)

rmse
mae
```
